### 2020-05-29 manuscript.Rmd
prefinal revision fpr submission

---
output:
  output: pdf_document
bibliography: bibliography.bib
csl: vancouver.csl
---

@@

*$^{1}$School of Health and Related Research, University of Sheffield, Sheffield, UK.;    
$^{2}$Advanced Wellbeing Research Centre, Sheffield Hallam University, Sheffield, UK.
$^{3}$The Centre for Sports Engineering Research, Sheffield Hallam University, Sheffield S9 3TU, UK*

@@

Target Journal: 1) Public Health (https://www.journals.elsevier.com/public-health)

@@

This study is a cross-sectional ecological analysis of the socio-economic disparities in geographic access to parkrun events in England at the end of 2018.

@@

We combined geo-location data on all English Lower Layer Super Output Areas and parkrun events to calculate geodesic distances to the nearest event for over 32,000 communities in England. We use this measure of geographic access to summarise the relationship between access and socio-economic deprivation, measured using the Index of Multiple Deprivation. We then used geographic coordinates of public green spaces in England to conduct a simple location-allocation analysis to identify 200 locations for future event locations that would maximize access.

@@

Insufficient physical activity is one of the leading causes of disease and disability worldwide.[@who2009] In the UK, around one in six deaths is attributable to low levels of physical activity.[@lee2012] It is also a major contributor to health inequalities, as people from low socio-economic backgrounds are both disproportionately likely to be inactive [@rowe2000sport; @farrell2014] and be affected by physical inactivity-related diseases [@petrovic2018contribution]. Increasing the physical activity levels of the population is therefore high on the public health agenda: it not only has the potential to improve quality of life, reduce mortality rates and alleviate the strain on health and social care services, but also reduce the gap in health inequalities [@vilhelmsson2018].

However, designing effective public health interventions that increase population physical activity is considerable challenge.[@reis2016] Implementing such interventions in a way that does not increase health inequalities might even be more difficult. Studies have shown that programmes to increase physical activity often fail to reach deprived communities and those most in need, suggesting access to physical activity facilities shows spatial and social inequality.[@white2009; @rideout2018] In this regard, parkrun, an international movement which organises free weekly 5 km running and walking events in public spaces, might provide valuable lessons.

Since its founding in 2004, as a small event in London with 13 participants, parkrun has grown to become one of the world's largest mass sporting events, with up to 360,000 participants in more than 20 countries.[@haake2018pr; @reece2019] The volunteer-led events are often characterised as accessible and inclusive.[@stevinson2013; @quirk2019] The organisation has been widely praised as being successful in encouraging participation particularly in individuals who were previously inactive.[@stevinson2019changes; @wiltshire2018]

Notwithstanding these subjective accounts, the expansion of parkrun in England, as elsewhere, has been largely grassroots, driven by *demand* rather than *need*. It might therefore be the case that parkrun events are primarily located in areas that are less deprived, while people living in more deprived communities may not have the same opportunities to participate. In 2019, Sport England announced funding to support the creation of 200 new parkrun events across England within three years, with the specific aim of increasing participation of individuals from lower socio-economic groups.[@reuters2018]

The aims of this study are two-fold: firstly, to evaluate whether geographic access to parkrun events in England is equitable across areas with different levels of deprivation. Secondly, to identify 200 optimal locations for future events in order to improve geographic access, in particular for deprived communities.

@@

This study is a cross-sectional ecological analysis of the socio-economic disparities in geographic access to parkrun events in England at the end of 2018. All analyses were conducted on the level of Lower Layer Super Output Areas (LSOAs), which divide England into 32,844 geographic units which, on average, have a population of approximately 1,700. We assessed the relationship between access, defined as the distance (as the crow flies) to the nearest parkrun event, and socio-economic deprivation, measured using the Index of Multiple Deprivation. In addition, we used information on public green spaces in England to conduct a simple location-allocation analysis, in order to identify 200 locations for future parkrun events that maximise access for the population.

@@

  1) For all 32,844 LSOAs, we retrieved geographic locations, defined by the coordinates of its population-weighted centroid, 2017 total population estimates, and the 2015 Index of Multiple Deprivation (IMD) from the Office for National Statistics.[@ons2011; @ons2015; ons2017]

  2) We included all 465 public parkrun events which were in operation in England by December 12th, 2018 - on this date, Sport England announced their plan to provide funding for setting up 200 additional parkrun events across England.[@parkrunGroup] The locations of the events were obtained from the parkrun UK website.[@parkrunweb]

  3) The locations of public green spaces in England were retrieved from an open dataset of Ordnance Survey.[@ordnance] parkrun events are held in various settings and terrains, and do not always require a single 5 km loop - some events have courses that involve running a combination of shorter loops.. After evaluating existing parkrun event courses, we decided to consider all public parks, gardens, and playing fields in England with an area of 0.1 km$^2$ or more potentially suitable for hosting events (n=2,842).

@@

  The socio-economic deprivation of LSOAs was measured using the 2015 IMD. It is a measure of relative deprivation, which has been used in many similar studies. The IMD combines 37 indicators from seven domains (income, employment, education and skills, health and disability, crime, housing and services, and living environment) into a single score. The score ranges from 0 (least deprived) to 100 (most deprived).[@smith2015]

@@

  More formally, the first step of the algorithm is defined by the following equation:

@@

  All data and the R source code that were used to generate the results of this study are provided on an open repository.[@githubrepo]

@@

  Ethical approval was obtained from the Sheffield Hallam University Ethics Committee (ER10776545). We did not collect any personal information, but only used aggregate secondary data. The parkrun Research Board approved this research project, and four of its members (AMB, HQ, EG, SSJH) were actively involved in the interpretation of findings and writing of this manuscript.

@@

  As of 12th December 2018, approximately 7%, 69%, and 91% of the English population lived within 1, 5, and 10 km distance of a parkrun event. Only 578,043 people (1% of the English population) lived more than 20 km from an event. The average (SD) and median (IQR) distance to the nearest parkrun event were 4.65 (4.22) and 3.39 (1.99-5.83). The largest distance was observed for the 2,259 people living on the Isles of Scilly, who live about 76 km away from the next parkrun event on the mainland. On average, each parkrun event is the closest event for 71 LSOAs (43), with a combined population of 119,612 (74,290). Further descriptive statistics are provided in table 1.

@@

![](./output/map_static.jpeg)
**Figure 1**: Map of England showing current parkrun events (blue circles) and recommended new event locations (red triangles) ranked in descending order of estimated effect on overall population-weighted access. Information on all 200 identified optimal green space locations are provided in the appendix.

@@

  The main finding, that geographic access to parkrun events is better in more deprived communities, is surprising. parkrun events are set up by volunteers, based on demand, not need. Studies have shown that the level of physical activity, and the availability of physical activity facilities generally declines with the level of deprivation.[@farrell2014socioeconomic] Opportunities for physical activity are often lacking in areas in most need.[e.g. @hillsdon2007] parkrun events, in contrast, seem to be often held in or near deprived areas, and are free to attend, giving anyone equal access, irrespective of their socio-economic background. Nevertheless, in a previous analysis, we found that participation in parkrun has a strong socioeconomic gradient with considerably higher participation rates in less deprived areas: about a third of all participants came from LSOAs in the least deprived quintile, while only 7% came from the most deprived quintile.[@smithschneider; @schneidersmith] This suggests that providing the opportunity to participate in parkrun events, while a necessary first step to enable participation, has not been sufficient to engage people living in deprived communities.[@white2009] This means, creating additional events in optimal locations could improve overall geographic access further, yet effective strategies will still be needed to increase equity in engagement in new and current events.

  There are several strengths of this study that deserve mention. Firstly, it is the first study of geographic access to parkrun in England - therefore the approach is novel and the data untapped. Secondly the analysis makes use of large and rich datasets, with over 30,000 LSOA and over 400 existing parkrun events it is unlikely that individual outliers are affecting the results. The almost universal availability of parkrun events throughout the country provides a learning opportunity to explore socio-spatial determinants influencing physical activity behaviour on a national scale.[@wiltshire2018] Our study contributes to the limited research in this area and identifies possible leads for further investigation.

  However, there are also limitations. Most importantly, geographic access is not measured as travel distance, or travel time, but as geodesic distance. In some cases, for example where natural barriers like hills, lakes or rivers block routes, the actual distance travelled may be far in excess of the geodesic distance.[@boscoe2012]

@@

  Finally. our analysis has been concerned only with determining to what extent deprived communities have geographic access to parkrun events. We did not investigate what other factors independently explain access more generally. It should be noted, however, that a contributing factor for the negative relationship between IMD and access is likely to be population density: deprived areas cluster in urban areas, where also most parkrun events take place. Rural areas, on the other hand, may therefore have worse geographic access. Further studies are required to better understand wider determinants of access to parkrun and/or physical activity facilities more generally.

@@

## Additional Information

@@

Elizabeth Goyder; Roles: Supervision, Conceptualization, Writing – Review & Editing

@@

```{r echo=FALSE}
  kable(read.csv("./output/map_legend.csv"),
        caption = "List of all 200 recommended green spaces for setting up parkrun events to maximise geographic access.",
        format = "latex",align = "lcccccc",
        col.names = linebreak(c("Park name",
                                "Total wtd. average distance",
                                "% Change from previous step",
                                "Park area (km2)",
                                "Longitude",
                                "Lattitude")),
        longtable = T, 
        booktabs = T) %>%
  kable_styling(font_size = 7,full_width = F) %>%
  column_spec(1, width = "3cm") %>%
  column_spec(2:4, width = "1cm") %>%
  column_spec(5:6, width = "1cm") 
```

>>>>

---
output:
  output: pdf_document
bibliography: bibliography.bib
csl: vancouver.csl
linestretch: 1.3
fontsize: 12pt
indent: false
---

@@

*$^{1}$School of Health and Related Research, University of Sheffield, Sheffield, UK.   
$^{2}$Advanced Wellbeing Research Centre, Sheffield Hallam University, Sheffield, UK.   
$^{3}$The Centre for Sports Engineering Research, Sheffield Hallam University, Sheffield S9 3TU, UK.*

@@

Target Journal: Public Health (https://www.journals.elsevier.com/public-health)

@@

This study is a cross-sectional ecological analysis of the socioeconomic disparities in geographic access to parkrun events in England at the end of 2018.

@@

We combined geo-location data on all English Lower Layer Super Output Areas and parkrun events to calculate geodesic distances to the nearest event for over 32,000 communities in England. We use this measure of geographic access to summarise the relationship between access and socioeconomic deprivation, measured using the Index of Multiple Deprivation. We then used geographic coordinates of public green spaces in England to conduct a simple location-allocation analysis to identify 200 locations for future event locations that would maximize access.

@@

Insufficient physical activity is one of the leading causes of disease and disability worldwide [@who2009]. In the UK, around one in six deaths is attributable to low levels of physical activity [@lee2012]. It is also a major contributor to health inequalities, as people from low socioeconomic backgrounds are both disproportionately likely to be inactive [@rowe2000sport; @farrell2014] and be affected by physical inactivity-related diseases [@petrovic2018contribution]. Increasing the physical activity levels of the population is therefore high on the public health agenda: it not only has the potential to improve quality of life, reduce mortality rates and alleviate the strain on health and social care services, but also reduce the gap in health inequalities [@vilhelmsson2018].

However, designing effective public health interventions that increase population physical activity is a considerable challenge [@reis2016; @kelly]. Implementing such interventions in a way that does not increase health inequalities might even be more difficult. Studies have shown that programmes to increase physical activity often fail to reach deprived communities and those most in need, suggesting access to physical activity facilities shows spatial and social inequality [@white2009; @rideout2018]. In this regard, parkrun, an international movement which organises free weekly 5 km running and walking events in public spaces, might provide valuable lessons.

Since its founding in 2004, as a small event in London with 13 participants, parkrun has grown to become one of the world's largest mass sporting events, with up to 360,000 participants in more than 20 countries [@haake2018pr; @reece2019]. The volunteer-led events are often characterised as accessible and inclusive [@stevinson2013; @quirk2019]. The organisation has been widely praised as being successful in encouraging participation particularly in individuals who were previously inactive [@stevinson2019changes; @wiltshire2018].

Notwithstanding these subjective accounts, the expansion of parkrun in England, as elsewhere, has been largely grassroots, driven by *demand* rather than *need*. It might therefore be the case that parkrun events are primarily located in areas that are less deprived, while people living in more deprived communities may not have the same opportunities to participate. In 2019, Sport England announced funding to support the creation of 200 new parkrun events across England within three years, with the specific aim of increasing participation of individuals from lower socioeconomic groups [@reuters2018].

The aims of this study are two-fold: firstly, to evaluate whether geographic access to parkrun events in England is equitable across areas with different levels of deprivation; and secondly, to identify 200 optimal locations for future events in order to improve geographic access, in particular for deprived communities.

@@

This study is a cross-sectional ecological analysis of the socioeconomic disparities in geographic access to parkrun events in England at the end of 2018. All analyses were conducted on the level of Lower Layer Super Output Areas (LSOAs), which divide England into 32,844 geographic units which, on average, have a population of approximately 1,700. We assessed the relationship between access, defined as the distance (as the crow flies) to the nearest parkrun event, and socioeconomic deprivation, measured using the Index of Multiple Deprivation. In addition, we used information on public green spaces in England to conduct a simple location-allocation analysis, in order to identify 200 locations for future parkrun events that maximise access for the population.

@@

  1) For all 32,844 LSOAs, we retrieved geographic locations, defined by the coordinates of its population-weighted centroid; 2017 total population estimates; and the 2015 Index of Multiple Deprivation (IMD) from the Office for National Statistics [@ons2011; @ons2015; @ons2017].

  2) We included all 465 public parkrun events which were in operation in England by December 12th, 2018 - on this date, Sport England announced their plan to provide funding for setting up 200 additional parkrun events across England [@parkrunGroup]. The locations of the events were obtained from the parkrun UK website [@parkrunweb].

  3) The locations of public green spaces in England were retrieved from an open dataset of Ordnance Survey [@ordnance]. parkrun events are held in various settings and terrains, and do not always require a single 5 km loop - some events have courses that involve running a combination of shorter loops. After evaluating existing parkrun event courses, we decided to consider all public parks, gardens, and playing fields in England with an area of 0.1 km$^2$ or more potentially suitable for hosting events (n=2,842).

@@

  The socioeconomic deprivation of LSOAs was measured using the 2015 IMD. It is a measure of relative deprivation, which has been used in many similar studies. The IMD combines 37 indicators from seven domains (income, employment, education and skills, health and disability, crime, housing and services, and living environment) into a single score. The score ranges from 0 (least deprived) to 100 (most deprived) [@smith2015].

@@

  More formally, the first step of the algorithm evaluates the following expression:

@@

  All data and the R source code that were used to generate the results of this study are provided on an open repository [@githubrepo].

@@

  Ethical approval was obtained from the Sheffield Hallam University Ethics Committee (ER10776545). We did not collect any personal information, and only used aggregate secondary data. The parkrun Research Board approved this research project, and four of its members (AMB, HQ, EG, SSJH) were actively involved in the interpretation of findings and writing of this manuscript.

@@

  As of 12th December 2018, approximately 7%, 69%, and 91% of the English population lived within 1, 5, and 10 km distance of a parkrun event respectively. Only 578,043 people (1% of the English population) lived more than 20 km from an event. The mean (SD) and median (IQR) distance to the nearest parkrun event were 4.65 (4.22) and 3.39 (1.99-5.83). The largest distance was observed for the 2,259 people living on the Isles of Scilly, who live about 76 km away from the next parkrun event on the mainland. On average, each parkrun event is the closest event for 71 LSOAs (43), with a combined population of 119,612 (74,290). Further descriptive statistics are provided in table 1.

@@

![](./output/map_static.jpeg)
**Figure 1: Map of England showing current parkrun events (blue circles) and recommended new event locations (red triangles) ranked in descending order of estimated effect on overall population-weighted access. Information on all 200 identified optimal green space locations are provided in the appendix.**

@@

  The main finding, that geographic access to parkrun events is better in more deprived communities, is surprising. parkrun events are set up by volunteers, based on demand, not need. Studies have shown that the level of physical activity, and the availability of physical activity facilities generally declines with the level of deprivation [@farrell2014socioeconomic] Opportunities for physical activity are often lacking in areas in most need [e.g. @hillsdon2007]. parkrun events, in contrast, seem to be often held in or near deprived areas, and are free to attend, giving anyone equal access, irrespective of their socioeconomic background. Nevertheless, in a previous analysis, we found that participation in parkrun has a strong socioeconomic gradient with considerably higher participation rates in less deprived areas: about a third of all participants came from LSOAs in the least deprived quintile, while only 7% came from the most deprived quintile [@smithschneider; @schneidersmith]. This suggests that providing the opportunity to participate in parkrun events, while a necessary first step to enable participation, has not been sufficient to engage people living in deprived communities [@white2009]. This means, creating additional events in optimal locations could improve overall geographic access further, yet effective strategies will still be needed to increase equity in engagement in new and current events.

  There are several strengths of this study that deserve mention. Firstly, it is the first study of geographic access to parkrun in England - therefore the approach is novel and the data untapped. Secondly the analysis makes use of large and rich datasets, with over 30,000 LSOA and over 400 existing parkrun events it is unlikely that individual outliers are affecting the results. The almost universal availability of parkrun events throughout the country provides a learning opportunity to explore socio-spatial determinants influencing physical activity behaviour on a national scale [@wiltshire2018]. Our study contributes to the limited research in this area and identifies possible leads for further investigation.

  However, there are also limitations. Most importantly, geographic access is not measured as travel distance, or travel time, but as geodesic distance. In some cases, for example where natural barriers like hills, lakes or rivers block routes, the actual distance travelled may be far in excess of the geodesic distance [@boscoe2012].

@@

  Finally, our analysis has been concerned only with determining to what extent deprived communities have geographic access to parkrun events. We did not investigate what other factors independently explain access more generally. It should be noted, however, that a contributing factor for the negative relationship between IMD and access is likely to be population density: deprived areas cluster in urban areas, where also most parkrun events take place. Rural areas, on the other hand, may therefore have worse geographic access. Further studies are required to better understand wider determinants of access to parkrun and/or physical activity facilities more generally.

@@

<!-- UNBLINDED:
## Additional Information

@@

Elizabeth Goyder; Roles: Supervision, Conceptualization, Writing – Review & Editing
-->

@@

```{r echo=FALSE}
  kable(read.csv("./output/map_legend.csv"),
        caption = "List of all 200 recommended green spaces for setting up parkrun events to maximise geographic access.",
        format = "latex",align = "lcccccc",
        col.names = linebreak(c("Park name",
                                "Total wtd. average distance",
                                "% Change from previous step",
                                "Park area (km2)",
                                "Longitude",
                                "Lattitude")),
        longtable = T, 
        booktabs = T) %>%
  kable_styling(font_size = 7,full_width = F) %>%
  column_spec(1, width = "4cm") %>%
  column_spec(2:4, width = "1.5cm") %>%
  column_spec(5:6, width = "1.5cm") 
```
